---
title: "Secondary Student Performance Prediction"
author: Xiaoyun Zhuang (xiaoyun6), Yu Huang (yuh4), Junjing Liu (junjing2)
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(tidyverse)
library(rpart)
library(rpart.plot)
library(caret)
```


```{r}
set.seed(42)
data["level"] = ifelse(data["G3"] >= 10, "pass", "fail")
trn_index = sample(1:nrow(data), round(nrow(data) * 0.9), replace = FALSE)
trn = data[trn_index, ]
tst = data[-trn_index, ]

rmse = function(act, pred) {
  sqrt(mean((act - pred) ^ 2))
}
mae = function(act, pred) {
  abs(mean(act - pred))
}
acc = function(act, pred) {
  mean(act == pred)
}
```

# Regression

## Without G1 and G2

```{r control}
fitcontrol = trainControl(method = "cv", number = 10)
```

```{r regression-w/o-G1-G2}
G3_form = G3 ~ . - G1 - G2 - level

# Linear regression
lm_mod = train(G3_form, data = data, method = "lm", trControl = fitcontrol, metric = "RMSE")
summary(lm_mod) 

lm_mod$results ## 2.743234

# tree model
tree_mod = train(G3_form, data = data, method = "rpart", trControl = fitcontrol, metric = "RMSE")
tree_mod$results ## 2.890101

# knn model
knn_mod = train(G3_form, data = data, method = "knn", trControl = fitcontrol, metric = "RMSE")
knn_mod$results ## 3.011992
```

```{r regression-w-G1-G2}
G3_wiz_form = G3 ~ . - level

# Linear regression
lm_mod2 = train(G3_wiz_form, data = data, method = "lm", trControl = fitcontrol, metric = "RMSE")
## ## G1 and G2 shows much more significance than other features
lm_G1G2 = train(G3 ~ G1 + G2, data = data, method = "lm", trControl = fitcontrol, metric = "RMSE")
lm_G1G2$results ## 1.224577	close to the full model >> that is to say, G1 and G2 are predictive enough for the final grade

sum((abs(predict(lm_G1G2, tst) - tst$G3) / 20) > 0.5)

summary(lm_mod2)
lm_mod2$results ## 1.279498

# tree model
tree_mod2 = train(G3_wiz_form, data = data, method = "rpart", trControl = fitcontrol, metric = "RMSE")
tree_mod2$results

# knn model
knn_mod2 = train(G3_wiz_form, data = data, method = "knn", trControl = fitcontrol, metric = "RMSE")
knn_mod2$results
```


# Appendix

## EDA

```{r}
# str(data)

Plot1 = data %>%
  ggplot(aes(x = school)) + geom_bar()

Plot2 = data %>%
  ggplot(aes(x = sex)) + geom_bar()

Plot3 = data %>%
  ggplot(aes(x = age)) + 
  geom_histogram(bins = 30)

Plot4 = data %>%
  ggplot(aes(x = address)) + geom_bar()

Plot5 = data %>%
  ggplot(aes(x = famsize)) + geom_bar()

Plot6 = data %>%
  ggplot(aes(x = Pstatus)) + geom_bar()
```


## Data Dictionary

- `sex`: student’s sex (binary: female or male)
- `age`: student’s age (numeric: from 15 to 22)
- `school`: student’s school (binary: **Gabriel Pereira** or **Mousinho da Silveira**)
- `address`: student’s home address type (binary: urban or rural)
- `Pstatus`: parent’s cohabitation status (binary: living together or apart)
- `Medu`: mother’s education (numeric: from 0 to 4[^a])
- `Mjob`: mother’s job (nominal[^b])
- `Fedu`: father’s education (numeric: from 0 to 4[^a])
- `Fjob`: father’s job (nominal[^b])
- `guardian`: student’s guardian (nominal: mother, father or other)
- `famsize`: family size (binary: ≤ 3 or > 3)
- `famrel`: quality of family relationships (numeric: from 1 – very bad to 5 – excellent)
- `reason`: reason to choose this school (nominal: close to home, school reputation, course preference or other)
- `traveltime`: home to school travel time (numeric: 1 – < 15 min., 2 – 15 to 30 min., 3 – 30 min. to 1 hour
or 4 – > 1 hour)
- `studytime`: weekly study time (numeric: 1 – < 2 hours, 2 – 2 to 5 hours, 3 – 5 to 10 hours or 4 – > 10 hours)
- `failures`: number of past class failures (numeric: $n$ if 1 ≤ $n$ < 3, else 4)
- `schoolsup`: extra educational school support (binary: yes or no)
- `famsup`: family educational support (binary: yes or no)
- `activities`: extra-curricular activities (binary: yes or no)
- `paidclass`: extra paid classes (binary: yes or no)
- `internet`: Internet access at home (binary: yes or no)
- `nursery`: attended nursery school (binary: yes or no)
- `higher`: wants to take higher education (binary: yes or no)
- `romantic`: with a romantic relationship (binary: yes or no)
- `freetime`: free time after school (numeric: from 1 – very low to 5 – very high)
- `goout`: going out with friends (numeric: from 1 – very low to 5 – very high)
- `Walc`: weekend alcohol consumption (numeric: from 1 – very low to 5 – very high)
- `Dalc`: workday alcohol consumption (numeric: from 1 – very low to 5 – very high)
- `health`: current health status (numeric: from 1 – very bad to 5 – very good)
- `absences`: number of school absences (numeric: from 0 to 93)
- `G1`: first period grade (numeric: from 0 to 20)
- `G2`: second period grade (numeric: from 0 to 20)
- `G3`: final grade (numeric: from 0 to 20)

[^a]: 0 – none, 1 – primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education
[^b]: teacher, health care related, civil services (e.g. administrative or police), at home or other.

## Links

[^1]: [Student Performance Data Set](http://archive.ics.uci.edu/ml/datasets/Student+Performance)
[^2]: [European Statistics](https://ec.europa.eu/eurostat/)
